Modeling memory and taxis to analyze migrations

### 2021-04-14 text/ms.Rmd
Updates on Model and manuscript

Diffusion-advection models have a long pedigree in animal movement modeling [@Skellam1951;@Turchin1998;Okubo2001].  Essentially, these models are grounded in the general idea that animal movements - somewhat like movements of physical particles - might combine a random (diffusive) component with a directed (advective) component and that the two can be separated and modeled.  While the links between diffusion models and movement data are somewhat tenuous (see @Gurarie2011a for quantitative links between obervations and diffusion parameters and @Ovaskainen2004 for applications to mark-recapture data), as a theoretical tool for exploring processes they are invaluable for their versatility and the relative ease of numeric computation of the partial differential equations (PDEs) that are used to describe diffusion-advection models mathematically. Even as equations become analytically intractable,

@@

But perhaps the greatest difference between animals and the randomly moving passive particles that are best described by diffusion models is the cognitive abilities, including social behavior and memory. Refinements to diffusion-advection equations have revealed conditions under which non-local information gathering [@Fagan2017] and behavioral switching may confer foraging advantages [@Fagan2019], in particular when resources are dynamic and patchy.   The interacting role of memory and sociality have been comparatively little studied.

Migration is a strategic behavior, typically involving a major trade-off between the costs of the migration against the benefits of accessing resources (whether food or predator refuge) that are highly seasonal and localized.  To perform a migration, a direct resource-following tactic (taxis) is likely to fail.

# Model

We study how a blending of a *tactical* (i.e. direct response to resource availability or perception) and a *strategic* (i.e. memory-driven and forward-thinking) behavior can help a forager navigate a dynamic and heterogeneous resource environment, with a particular emphasis on the aquisition and adaptabilty of **migratory behavior** in a dynamic, seasonal resource environment, i.e. one in which the distribution of the resource at a given time is approximately equal to that resource distribution at a fixed previous time period, i.e. $h(t) \approx h(t - \tau)$.

Our model is an advection diffusion with two advective terms corresponding to a direct response to the environment and to memory. The *tactical* process is straightforward taxis up a resource gradient, while the *collective memory* corresponds to a tendency of the organisms to repeat the behavior of the population in previous years.  Thus:

@@

where the variables $u(x,t)$ represents a population distributed in time and space, $h(x,t)$ is the distribution of the resource, $m(x,t,y)$ is the memory component in space and time at (discrete) year $y$, $\varepsilon$ is a rate of randomness (diffusion), $\alpha$ is the strength of advection in the direction of the resource $h(x,t)$ gradient), and $\beta$ is the strength of advection up the gradient memory.  The memory component we model as a mixture of a reference (long-term) memory, modified by more recent experiences as follows:
$$m(x,t) = \kappa^y \, u(x, t - y\tau) + \left(1-\kappa^y\right)\,u(x, t-\tau)\,\,\,\,(2)$$
Here, $y \in \{0,1,2,...\}$ represents the count of a given year, i.e. $y = \lfloor t/\tau \rfloor$. Thus, $u(x, t - y\tau)$ is the initial (year 0) "learned" collective behavior, $u(x, t-\tau)$ is the preceding year's collective behavior and $\kappa \in (0,1)$ is the relative strenfth of the long-term (reference) memory.   When $\kappa = 1$, all of the memory is reference memory, and the system essentially responds to it's initial learned behavior as if it were a resource. If $\kappa = 0$, the memory is short-term, and the process responds only to the behavior from the previous year.  With $0 < \kappa < 1$, the reference memory fades with time.

For reference and replicability, I include the incrementally constructed R code, which relies heavily on the `deSolve` [@Soetaert2010] and `ReacTran` [@Soetaert2012] packages, which efficiently implement transport ODE and PDE systems in R.

```{r, message=FALSE, warning=FALSE}
require("memorymigration")
```

@@

```{r}
ForagingMemoryModel <- function(t, pop, parms, pop_lag, resource, 
                                dx = attributes(pop)$dx){
  tran.1D(C = pop, D = parms["epsilon"], 
                   flux.up = 0, flux.down = 0, 
                   v = parms["alpha"] * diff(resource)/dx + 
                       parms["beta"] * diff(pop_lag)/dx, 
                   dx = dx)
}
```

As a simple illustration of a model run, below the ultimate evolution of a forager that begins in the left part of space and uses a combination of the previous year's location (to the right) and the current resource distribution, and ultimately "splits up" into being attracted to the resource in the middle and its previous year's distribution, with a somewhat stronger taxis towards the resource than the memory:

```{r GradientExample, echo=FALSE, cache=TRUE, message=FALSE, warning=FALSE}
dx <- .01
X.max <- 100
x.edges <- seq(0,100, length=X.max/dx + 1)
x.mids <- (x.edges[-1] + x.edges[-length(x.edges)])/2

parameters <- c(epsilon = 0.1, alpha = 4, beta = 2)
pop <- dnorm(x.mids, mean=20, sd = 2) %>% {./sum(.)/dx} 
attributes(pop)$dx <- dx

pop_lag <- c(0,dnorm(x.mids, mean=80, sd = 2) %>% {./sum(.)/dx},0)
resource <- c(0,dnorm(x.mids, mean = 50, sd = 4) %>% {./sum(.)/dx},0)

pop.inf <- steady.1D(func = ForagingMemoryModel, 
                   y = pop, parms = parameters, 
                   resource = resource, 
                   pop_lag = pop_lag, 
                   nspec=1, 
                   dx = attributes(pop)$dx)$y %>% {./sum(.)/attributes(pop)$dx}

n <- length(x.mids)

cols <- c("black", "darkgrey", "green",  "red")
lwds <- c(1,3,3,1)
ltys = c(1,3,3,1)

par(bty = "l", mgp = c(1.5,.25,0), tck = 0.01, las = 1)

plot(x.mids, pop, type = "l", xlab = "space", ylab = "density", 
     col = cols[1], lwd = lwds[1], lty = ltys[1])
lines(x.mids, pop_lag[-c(1,n+1)], col = cols[2], lwd = lwds[2], lty = ltys[2])
lines(x.mids, resource[-c(1,n+1)], col = cols[3], lwd = lwds[3], lty = ltys[3])
lines(x.mids, pop.inf, col = cols[4], lwd = lwds[4], lty = ltys[4])

legend("top", legend = c("pop. initial", "pop. previous year", "resource", "pop. final"), 
       lty = ltys, lwd = lwds, col = cols, bty = "n")
#title(parse(text = printParameters(p)))

printParameters <- function(p) 
  parse(text = paste0("list(",paste(names(p), p, sep = "==", collapse = ", "),")"))

title(printParameters(parameters))

```

This example is completely static and symmetric around the two drivers.  Note that all the populations (previous, initial, final) are normalizes (integrate to 1) and that the space is defined between 0 and 100.

# Dynamic initial population

In the following example, we let the population be "initialized" via a sinusoidal migration pattern, while the resource is fixed in the middle of the space (as above).  The memory is a one-year memory.  In this context, migration is not, in fact, optimal.  But the important piece is to see how the response mixes attraction to the resource with attraction to the previous year's behavior.  Note that the year here (for simplicity, speed and symmetry), is 100 days long.

```{r setupEnvironment}
getSinePop <- function(tau, X.min = 0, X.max = 100, dx, 
                       peak.max, peak.min, sd){
  Time <- 1:tau

  X.edge <- seq(X.min, X.max, dx)
  X <- (X.edge[-1] + X.edge[-length(X.edge)])/2
  pop <- outer(Time, X,
                     function(t,x) 
                       dnorm(x, mean = X.max/2 + (peak.max-peak.min) / 2 * sin( 2*(t * pi)/max(Time)),sd = sd))
  pop <- apply(pop, 1, function(x) x/sum(x)/dx) %>% t
  list(pop = pop, X = X, time = Time, dx = dx, tau = tau)
}

world <- getSinePop(tau = 100, X.min = 0, X.max = 100, dx=1, 
                    peak.max = 80, peak.min = 20, sd = 10)

fr1 <- dnorm(world$X, mean = 80, sd = 10) %>% {./sum(.)/world$dx}
fr2 <- dnorm(world$X, mean = 20, sd = 10) %>% {./sum(.)/world$dx}
fixed.Resource <- matrix((fr1 + fr2)/2, byrow = TRUE, 
                         nrow = nrow(world$pop), 
                         ncol = ncol(world$pop))

world$resource <- fixed.Resource
```

The following examples illustrate how an initially migrating population evolves, even as it is attracted to a resource that is distributed in two different locations. In these examples, the first year (year 0) is a sinusoidal migration.  We run the simulations for 3 subsequent years.

## Only memory

```{r memoryOnly, fig.height = 5, fig.width = 10}

parameters <- c(epsilon = 5, alpha = 0, beta0 = 20, beta1 = 5e2)

eval <- FALSE
if(eval){
  MemSim <- runManyYears(world, Parameters = parameters, 
                         n.years = 3, 0.99)
  save(MemSim, file = "../results/MemSim.rda")
} else load("../results/MemSim.rda")

par(mfrow = c(1,2), oma = c(0,0,3,0))
plotYearList(MemSim, world, tau = tau)
plotYearList(MemSim, world, tau = tau, persp = TRUE, scale = FALSE, expand = 100)
title(printParameters(parameters), outer = TRUE)
```

The population roughly echoes its initial track, but degrades from year to year.

## Resource only

In the following simulations, we add a resource distributed as follows in two concentrations:

```{r fixedResource, fig.height = 3, fig.width = 4}
with(world, 
  image(X, 1:tau, resource, main = "resource", ylab = "x")
)
```

```{r resourceOnly, fig.height = 5, fig.width = 10}
parameters <- c(epsilon = 5, alpha = 5e2, beta0 = 0, beta1 = 0)

eval <- FALSE
if(eval){
  FixedResourceSim <- runManyYears(world, Parameters = parameters, n.years = 3, 0.99)
  save(FixedResourceSim, file = "../results/FixedResourceSim.rda")
} else load("../results/FixedResourceSim.rda")

par(mfrow = c(1,2), oma = c(0,0,3,0))
plotYearList(FixedResourceSim, world, tau = tau)
plotYearList(FixedResourceSim, world, tau = tau, persp = TRUE, scale = FALSE, expand = 100)
title(printParameters(parameters), outer = TRUE)
```

Here, the population disregards its memory and only pursues a tactic stategy. Within a year, the population has gathered around the resource.

## Resource and memory

Finally, we combine both resource following and memory:

```{r resourceAndMemory, fig.height = 5, fig.width = 10}
parameters <- c(epsilon = 5, alpha = 3e2, beta0=20, beta1 = 5e2)

eval <- FALSE
if(eval){
  MemResourceSim <- runManyYears(world, Parameters = parameters, n.years = 3, 0.99)
  save(MemResourceSim, file = "../results/MemResourceSim.rda")
} else load("../results/MemResourceSim.rda")

par(mfrow = c(1,2), oma = c(0,0,3,0))
plotYearList(MemResourceSim, world, tau = tau)
plotYearList(MemResourceSim, world, tau = tau, persp = TRUE, scale = FALSE, expand = 100)
title(printParameters(parameters), outer = TRUE)
```

Here, the strong cultural memory is slowly blended with a strategy to simply concentrate in one of the two resource hot-spots.

# Seasonal resource

There is obviously nothing particularly adaptive about migrating when resources are fixed.  In the following scenaries, we explore a seasonal resource. We developed (with some fuss!) a seasonal resource function with the following properties:

1. The total amount of resource across space is constant throughout the year.
2. At the beginning, middle, and end of the year the resource is uniformly distributed. 
3. At some peak time $t_r < \tau/2$, the resource concentrates at a location $x_r < \chi/2$ with a spatial deviation $\sigma_x$ and a temporal deviation $\sigma_t$ (where $\tau$ is the length of the year and $\chi$ is the extent of the spatial domain).
4. The resource peaks exactly symmetrically at time $\tau - t_r$ and location $\chi - x_r$ with the same variance $\sigma_r$.

To generate a resource with these properties, we distributing the resource in space as a beta distribution, where the two shape and scale parameters vary sinusoidally in such a way as to fulfill the criteria above.  Thus:

$$R(x,t, \theta) = \chi B(x/\chi, a(t, \theta), b(t, \theta))$$

where $\chi$ is the maximum value (domain) of $x$, $B(x, a, b)$ is the beta distribution, $\theta$ represents the set of parameters $t_r, x_r, \sigma_t, \sigma_x$, and the two shape parameters are given by:

$$a(t) = \frac{m}{s^2}( s^2 + m - m^2)$$
$$b(t, x', \sigma') = (m-1)\left(1 + \frac{m}{s}(m-1)\right) $$

where $m(t)$ and $s(t)$ describe the dynamic mean and variance of the resource peak.  These equations are solutions to the mean and variance of the beta distribution, $\mu = \alpha/(\alpha + \beta)$, $\sigma^2 = \frac{\alpha\beta}{(\alpha+\beta)^2(\alpha+\beta+1)}$.

The means and variances themselves are Gaussian pulses, with the mean peaking at $x_r$ at time $t_r$ with standard deviation $\sigma_t$ and at $\chi - x_t$ at time $\tau - t_r$ and the standard deviation pulsing from $\chi/\sqrt{12}$ (corresponding to a uniform distribution) at times 0, $\tau/2$ and $\tau$ down to $\sigma_x$ at $t_r$ and $\tau - t_r$, with standard deviation (in time) $\sigma_t$.

All of these details boil down to the ability to generate various seasonal scenarios.

```{r exampleResourcePlots, fig.width = 10, fig.height = 6}
source("functions2.R")

world.R1 <- world %>% 
  list_modify(resource = getPulsedResource(1:world$tau, world$X, c(t.peak = 25, t.sd = 9, x.peak = 80, x.sd = 9)))

world.R2 <- world %>% 
  list_modify(resource = getPulsedResource(1:world$tau, world$X, c(t.peak = 25, t.sd = 9, x.peak = 80, x.sd = 5)))

world.R3 <- world %>% 
  list_modify(resource = getPulsedResource(1:world$tau, world$X, c(t.peak = 25, t.sd = 5, x.peak = 80, x.sd = 9)))

world.R4 <- world %>% 
  list_modify(resource = getPulsedResource(1:world$tau, world$X, c(t.peak = 25, t.sd = 5, x.peak = 80, x.sd = 5)))

par(mfrow = c(2,2), mar = c(3,3,3,1), mgp = c(2,.25,0), las = 1, tck = 0.01)
with(world.R1, {     
  image(1:tau,X,resource, main = "Broad, long peak")
  contour(1:tau,X,resource, add = TRUE)
  })
with(world.R2, {     
  image(1:tau,X,resource, main = "Narrow, long peak")
  contour(1:tau,X,resource, add = TRUE)
  })
with(world.R3, {     
  image(1:tau,X,resource, main = "Broad, short peak")
  contour(1:tau,X,resource, add = TRUE)
  })
with(world.R4, {     
  image(1:tau,X,resource, main = "Narrow, short peak")
  contour(1:tau,X,resource, add = TRUE)
  })

```

# Running the model in a seasonal environment
# Note: to view any of the resource models, change the world

In these pulsed resources, we might assume that a *combination* of memory and tactic responses would yield more optimal behavior, and might observe some shift in the migratory strategy.

## Memory only

```{r PulseSim1, fig.width = 10}
parameters <- c(epsilon = 5, alpha = 0, beta0 = 20, beta1 = 5e2)
Memory_PulsedResource_Sim <- runManyYears(world.R3, Parameters = parameters, n.years = 6, 0.99)

par(mfrow = c(1,2), oma = c(0,0,3,0))
plotYearList(Memory_PulsedResource_Sim, world.R3, tau = tau)
plotYearList(Memory_PulsedResource_Sim, world.R3, tau = tau, persp = TRUE, scale = FALSE, expand = 100)
title(printParameters(parameters), outer = TRUE)
```

## Tactics only

This model does a surprisingly good job aggregating at the appropriate resource, but appears to only locally aggregate, not to migrate per se.

```{r PulseSim2, fig.width = 10}
parameters <- c(epsilon = 2, alpha = 4e2, beta0 = 0, beta1 = 0)
Tactic_PulsedResource_Sim <- runManyYears(world.R1, Parameters = parameters, n.years = 6, 0.99)

par(mfrow = c(1,2), oma = c(0,0,3,0))
plotYearList(Tactic_PulsedResource_Sim, world.R1, tau = tau)
plotYearList(Tactic_PulsedResource_Sim, world.R1, tau = tau, persp = TRUE, scale = FALSE, expand = 100)
title(printParameters(parameters), outer = TRUE)
```

## Memory and tactics

It is not clear that the hybrid model recovers a true migration, but it seems to retain somewhat more concentration along the migratory corridors.

```{r PulseSim3, fig.width = 10}
parameters <- c(epsilon = 5, alpha = 2e2, beta0 = 20, beta1 = 3e2)
Hybrid_PulsedResource_Sim <- runManyYears(world.R1, Parameters = parameters, n.years = 6, 0.99)

par(mfrow = c(1,2), oma = c(0,0,3,0))
plotYearList(Hybrid_PulsedResource_Sim, world.R1, tau = tau)
plotYearList(Hybrid_PulsedResource_Sim, world.R1, tau = tau, 
             persp = TRUE, scale = FALSE, expand = 100)
title(printParameters(parameters), outer = TRUE)
```

## Memory and tactics: Narrow resource

Here we compare a memory with and without memory exploiting the "narrow - long" resource combination, and the tendency to

### No memory - strong tactics

Notable concentrations at the resource peaks, but a lot of diffuse behavior in the interim seasons

```{r PulseSim4_tactic, fig.width = 10}
parameters <- c(epsilon = 2, alpha = 2e2, beta0 = 0, beta1 = 0)
Tactic_PulsedResource_Sim2 <- runManyYears(world.R1, Parameters = parameters, n.years = 6, 0.99)

par(mfrow = c(1,2), oma = c(0,0,3,0))
plotYearList(Hybrid_PulsedResource_Sim2, world.R1, tau = tau)
plotYearList(Hybrid_PulsedResource_Sim2, world.R1, tau = tau, persp = TRUE, scale = FALSE, expand = 100)
title(printParameters(parameters), outer = TRUE)
```

### Hybrid model

Memory appears to reinforce the aggregations at the (narrow) resource peaks

```{r PulseSim4_memory, fig.width = 10}
parameters <- c(epsilon = 5, alpha = 2e2, beta0 = 20, beta1 = 3e2)
Hybrid_PulsedResource_Sim3 <- runManyYears(world.R4, Parameters = parameters, n.years = 6, 0.99)

par(mfrow = c(1,2), oma = c(0,0,3,0))
plotYearList(Hybrid_PulsedResource_Sim3, world.R4, tau = tau)
plotYearList(Hybrid_PulsedResource_Sim3, world.R4, tau = tau, persp = TRUE, scale = FALSE, expand = 100)
title(printParameters(parameters), outer = TRUE)
```

## Example
We apply the three indices to an example model. In this model, we let the population be "initialized" via a sinusoidal migration pattern, while the resource is seasonal in a broad, long peak. The model uses the parameters $\epsilon = 5, \alpha = 300, \beta0=100, \beta1 = 300$ so that the migratory population combines both memory and a tactical strategy. We run the model for 12 years and compute the three indices. Then we plot the three indices to the view the trend of the migratory population over time.

```{r}
dx <- 1

parameters <- c(epsilon = 5, alpha = 300, beta0=100, beta1 = 300)
world.R1 <- getSinePop(tau = 100, X.min = 0, X.max = 100, dx=2, 
                    peak.max = 80, peak.min = 20, sd = 10)
world.R1$resource <- getPulsedResource(world.R1$time, world.R1$X, c(t.peak = 25, t.sd = 9, 
                                                           x.peak = 80, x.sd = 9))

#run simulation for 12 years 
M <- runManyYears(world.R1, Parameters = parameters, n.years = 6, 1)

world.R2 <- world.R1 %>% list_modify(pop = M$Year6)
M2 <- runManyYears(world.R2, Parameters = parameters, n.years = 6, 1)

  indices1 <- sapply(M, computeIndices, resource = world.R1$resource, world = world.R1) %>% t
  indices2 <- sapply(M2, computeIndices, resource = world.R2$resource, world = world.R2) %>% t
  I <- rbind(indices1, indices2)

  matplot(I, type="o", pch = 19, lty = 1) #year 6 and year 7 are the same so it is plotted twice 
  legend("topright", pch = 19, col = 1:3, legend = colnames(I))

```
Then we run the model to stop running the simulation at a certain threshold when the population reaches a "dynamic equilibrium state". In this example, the model only runs three years before hitting the threshold. 
```{r}
#simulation runs for 3 years since that is when threshold is hit

M <- runManyYears(world.R1, Parameters = parameters, n.years = 6, 0.99)
plotYearList(M, world.R1, tau = tau)

computeCohesiveness(M[[3]], world.R1)["SC.mean"]
computeMigratoriness(M[[3]], world.R1)$overlap
computeEfficiency(M[[3]], world.R1$resource, world.R1)
computeIndices(M[[3]], world.R1$resource, world.R1)

```

>>>>

Diffusion-advection models have a long pedigree in animal movement modeling [@Skellam1951;@Turchin1998;@Okubo2001].  Essentially, these models are grounded in the general idea that animal movements - somewhat like movements of physical particles - might combine a random (diffusive) component with a directed (advective) component and that the two can be separated and modeled. While direct relationships between diffusion models and movement data are somewhat tenuous [@Gurarie2011a], as a theoretical tool for exploring processes they are invaluable for their versatility and the relative ease of numeric computation of the partial differential equations (PDEs) that are used to describe diffusion-advection models mathematically. Even as equations analytically intractable,

@@

However, perhaps the greatest difference between animals and the randomly moving passive particles that are best described by diffusion models is the cognitive abilities, including social behavior and memory. Refinements to diffusion-advection equations have revealed conditions under which non-local information gathering [@Fagan2017] and behavioral switching may confer foraging advantages [@Fagan2019], in particular when resources are dynamic and patchy. The interacting role of memory and sociality have been comparatively little studied.

Migration is a strategic behavior, typically involving a major trade-off between the costs of the migration against the benefits of accessing resources (whether food or predator refuge) that are highly seasonal and localized (**refs**).  While some migrants - both terrestrial and avian - follow a "green wave" of spring vegetation as it flowers across altitudinal or latitudinal gradients [@Bischof2012;@Kolzsch2015], many others "jump" the green wave, anticipating that critical resources will be available at the time of arrival (**refs**).

We anticiapted that under many conditons a blending of *tactical* (i.e. direct response to resource availability or perception) and *strategic* (i.e. memory-driven and forward-thinking) behavior can help a forager navigate a dynamic and heterogeneous resource environment, with a particular emphasis on the aquisition and adaptabilty of **migratory behavior** in a dynamic, seasonal resource environment To explore the interaction of memory and resource following as an adaptive strategy in dynamic environments. Specifically, we explore a one-dimensional advection diffusion model with long-term and short-term memory, and run the model in a dynamic, seasonal environment.

# Methods

## Memory movement model

Our model is an advection diffusion with two advective terms corresponding to a direct response to the environment and to memory. The *tactical* process is positive response to a resource gradient, while the *collective memory* corresponds to a tendency of the organisms to repeat the behavior of the population in previous years.  Thus:

@@

where the variables $u(x,t)$ represents a population distributed in time and space, $h(x,t)$ is the distribution of the resource, $m(x,t,y)$ is the memory component in space and time at (discrete) year $y$, $\varepsilon$ is a rate of randomness (diffusion), $\alpha$ is the strength of advection in the direction of the resource $h(x,t)$ gradient), and $\beta$ is the strength of advection up the gradient memory.  The memory component we model as a mixture of a reference (long-term) memory, modified by more recent experiences as follows:
$$m(x,t) = \kappa^y \, u(x, t - y\tau) + \left(1-\kappa^y\right)\,u(x, t-\tau)\,\,\,\,(2)$$
Here, $y \in \{0,1,2,...\}$ represents the count of a given year, i.e. $y = \lfloor t/\tau \rfloor$. Thus, $u(x, t - y\tau)$ is the initial (year 0) "learned" collective behavior, $u(x, t-\tau)$ is the preceding year's collective behavior and $\kappa \in (0,1)$ is the relative contribution of the long-term (reference) memory.   When $\kappa = 1$, all of the memory is reference memory, and the system essentially responds to it's initial learned behavior as if it were a resource. If $\kappa = 0$, the memory is entirely short-term, and the process responds only to the behavior from the previous year.  With $0 < \kappa < 1$, the reference memory fades with time.  The model is confined to a one-dimensional bounded domain $[0,\chi]$, with no flux outside of the boundaries i.e. ${\frac{\partial u(0,t)}{\partial t}} = {\frac{\partial u(\chi,t)}{\partial t}} =0$,  Thus the total population is fixed at all times and found within the domain.

A few special cases of this model have been studied in the past. @Shigesada1979 (section 3) thoroughly analyzed the resource following diffusion model with no memory and a static resource (i.e.~$\beta = 0$ and $h(x,t) = h(x)$), showing that at equilibrium, the log of the population density is proportional to $h(x)/\alpha$.  @Shi2019 consider a single period memory model ($\kappa = 0$) in detail, with the addition of an independent birth-death process. Importantly, the authors prove the wellposedness (existence, uniqueness, and positivity) of this model, with arguments that are equally valid for our model. @Fagan2017 numerically explored an extension of this model with no memory but a dynamic resource and the addition of non-local information, i.e.~a model where the resource $h(x,t)$ is "perceived" via a convolution with a detection function, while @Fagan2019 added a behavioral switching behavior to the non-local information model.  This study

## Environment and seasonal resource

We ran our numerical experiments on a spatial domain ranging from 0 to $\chi = 100$, and a period of duration $\tau = 100$ (i.e.~a 100 day "year").  We are interested in an approximately periodic resource dynamic, i.e. one in which $h(x,t \approx h(x, t-\tau))$ and $\tau$ is the period of the process. to do this.  Furthermore, we were mainly interested in a "non-surfable" resource where the patches are isolated and pulse periodically.

Thus, the resource model consists of two bivariate normal pulses centered at $\mu_p$ at $t_p$ and $\chi - \mu_p$ at $\tau - t_p$ (e.g. location 80 and 20 at times 25 and 75, respectively). These pulses have a shared time scale of duration $\sigma_t$ and a spatial scale of extent $\sigma_x$, the standard deviation in the time and space dimension respectively.  This flexible formulation allows us to compare long (duration) and narrow (spatially), long and wide, short and narrow, and short and wide pulses, and to easily simulate scenarios where there is drift or randomness in the location or timing of the pulses. Note that in order to make the magnitude of the resource pulse comparable to the magnitude of the memory, which integrates to 1 at every time point, we normalize the resource such that its average value across the year is also 1.

## Metrics

The main metrics we are interested are *migratoriness* and *foraging efficiency*.  The migratoriness is measured as the maximum distance between the median of the population distribution within a given year.  The median is a more reliable metric than the mean, as even as distributions become highly assymetric - pulling the mean in one directon or another - the median can  remain consistently in the middle of the spatial domain indicating a net flux of animals of 0. This is the anticipated (and realized) result of a purely "tactical" model pursuing the periodic resource pulses.

To quantify the foraging efficiency, i.e., the organisms’ ability to track the distribution of the resources over space and time, we use a continuous form of the Bhattacharyya Coefficient [@Bhattacharyya1943] which quantifies the similarity between two distributions.  We compute this coefficient at every time point in a given year, and take the mean across the equilibrium year to determine foraging efficiency (FE). Thus, the foraging efficiency index is: 
$$FE = {1\over \tau} \int_{0}^\tau \int_{0}^{\chi} \sqrt{u(x,t) \, h(x,t)} \,\, dx\,dt$$
where the spatial integral is taken over the domain.  This metric is constraned to be between 0 and 1.

Simulations are run until a (quasi)-equilibrium state is achieved, i.e. where the Bhattacharya index across subseqyuent years reaches a value of 0.9999.  Once that state is attained, we compute the metrics. We additionally collect the number of runs it

## Simulation studies

We numerically solved this model using numerical differencing of a system of ordinary differential equations (ODE's) approximating the PDE in equation (2) with the Runge-Kutte algorithm using the `deSolve` [@Soetaert2010] and `ReacTran` [@Soetaert2012] packages in R.

We assessed a wide range of parameter values and resource geometries and dynamics with the goal of answering the three main questions.

Q1. What

@@

![](figures/ScanningResourceMemoryRatios_part1.png)

![](figures/FinalMigratoryFollowing.png){width=50%}
